Supplementary Information

### 2020-08-08 doc/manuscript/MEE_manuscript.Rmd
Update MEE_manuscript.Rmd
removed cloud group from manuscript and schematic
#59

Monthly climate data for 1901-20XX were obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020]. 
Variables considered here included mean, minimum, and maximum temperatures; daily temperature range; precipitation; wet day frequency; cloud cover; and potential evapotranspiration (PET). In the CRU database, gaps are filled with monthly means... (**[ISSUE #45 in ForestGEO_climate-senstivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/45): handling gaps**).
(*Describe criteria for excluding variables/ time frames.*)
(*Describe BCI data.* )

@@

![**Figure 1 | DRAFT Schematic illustrating our analysis process.** This analysis is conducted separately for each site. (*CO2 needs to be removed from this figure.*) ](tables_figures/schematic_figure.png)

@@

First, we identified the most important climate driver for each of three categories of variables: a temperature group (mean, min, and max temperature; PET), a precipitation group (precipitation, number of days with precipitation), and a group linked to cloud cover and temperature variability ("cloud" group; daily temperature range, percent cloud cover, and again PET). 
Because the variables in the cloud group are linked to both temperature and precipitation, this group was given lowest precedence in the variable selection process.
These groups were defined based on the dependencies of these variables on one another in the CRU database [@harris_updated_2014; @harris_version_2020], as well as climatological and biological considerations.
Our first step was detrending to remove the influence of all non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), which is essential for identifying climatic drivers (DENDRO_REFS).
Specifically, we used a generalized least squares (GLS) model to fit a spline to individual tree growth records ($log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$), thereby producing residuals.
We note that an an attempt at the analysis without detrending failed to produce results consistent with biological expectations and results obtained via traditional dendrochronological methods, and generally failed to identify significant climate drivers.

We then used the *climwin* package in R (REF) [@van_de_pol_identifying_2016] to identify the most significant climate driver and time window for each of the three climate variable groups. 
*Climwin* searches through multiple climate variables and over a wide range of time frames to identify which are most strongly correlated to the biological variable of interest--in this case, residuals of $log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$.
We began by verifying that *climwin* identified similar climate variable-month combinations as what would be identified using traditional methods, as detailed in Appendix S2. (**[ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35))

Within *climwin*, we specified a mixed effects model using species and tree identity as random effects: `residual ~ [climate] + (1 | sp) + (1 | treeID)`. 
Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of time periods, at monthly resolution, starting in the previous May and ending in August of the current year.
We note that analyzing all species together yields the most significant climate drivers across the full set of cores from each site (our goal here), whereas identification of the top climate drivers for individual species would be optimized by analyzing each species separately.
*Climwin* can check for linear and quadratic relationships, and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@van_de_pol_identifying_2016]. 
We specified quadratic relationships only because (i) quadratic relationships are more consistent with known biological mechanisms (see Introduction), (ii) preliminary tests revealed that quadratic fits usually had lower AIC, and when not there tended to be little difference in AIC and the curve would approximate a straight line, (iii) modeling only quadratic improves coding efficiency and speeds up the analysis process.
For each group of candidate climate variables, we move forward with the best variable over the time window identified by climwin as a candidate climate variable for the multivariate models. 
If PET came out as the best variable in either the temperature or cloud group and there was any overlap in the time windows of variables from these groups, the cloud group variable was dropped from the analysis on the logic that it the top variables for these groups were too closely related.

@@

**The most influential climate drivers...[what tended to be selected?].**
- Time windows
  - often current growing season
  - some exceptions
- Selection within groups:
  - precip vs n precip days
  - temperature group: mean, min, max or PET?
  - t range group: which tended to be most important?

![**Figure 3 | Climate senstivity for all sites.** Columns include the top variables in each grouping, with PET placed in the temperature group if included in the model. For each species, relationships are plotted if included in top model. Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD. (IN-PREP FIGURE.) ](tables_figures/pre_temp_cld_groups.png)

@@

**Non-linear responses to climate variables were common (Fig. 3).**
- Growth almost always increased with precipitation up to the long-term mean, but often declined under high precipitation.
- The most common response to temperature was unimodal, often peaking within 1SD of long-term mean. (**depending on how strongly this comes out, may warrent separate paper**)
- (results for cloudiness group)

@@

![**Figure 4 | Growth sensitivity to DBH: (a) $\Delta r$, (b) $BAI$, (c) $\Delta AGB$.** Only signficant relationships are shown. It's tough to fit all the plots, so we may have to adjust this somehow.](tables_figures/DBH_responses.png)

>>>>

Monthly climate data for 1901-20XX were obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020]. 
Variables considered here included mean, minimum, and maximum temperatures; potential evapotranspiration (PET); precipitation; and wet day frequency. In the CRU database, gaps are filled with monthly means... (**[ISSUE #45 in ForestGEO_climate-senstivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/45): handling gaps**).
(*Describe criteria for excluding variables/ time frames.*)
(*Describe BCI data.* )

@@

![**Figure 1 | DRAFT Schematic illustrating our analysis process.** This analysis is conducted separately for each site.](tables_figures/schematic_figure.png)

@@

First, we identified the most important climate driver for each of two categories of variables, all at the monthly timescale: a temperature group (mean, min, and max temperature; PET) and a precipitation group (precipitation, number of days with precipitation). 
Our first step was detrending to remove the influence of all non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), which is essential for identifying climatic drivers (DENDRO_REFS).
Specifically, we used a generalized least squares (GLS) model to fit a spline to individual tree growth records ($log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$), thereby producing residuals.
We then used the *climwin* package in R (REF) [@van_de_pol_identifying_2016] to identify the most significant climate driver and time window for each variable group. 
*Climwin* searches through multiple climate variables and over a wide range of time frames to identify which are most strongly correlated to the biological variable of interest--in this case, residuals of $log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$.
We began by verifying that *climwin* identified similar climate variable-month combinations as what would be identified using traditional methods for individual species, as detailed in Appendix S2. (**[ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35))

We then used *climwin* to identify the best climate variables across the full set of cores from each site, noting that whereas our goal was to identify the most significant climate drivers at a site level, identification of the top climate drivers for individual species would be optimized by analyzing each species separately.
Within *climwin*, we specified a mixed effects model using species and tree identity as random effects: `residual ~ [climate] + (1 | sp) + (1 | treeID)`. 
Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of time periods, at monthly resolution, starting in the previous May and ending in August of the current year.
*Climwin* can check for linear and quadratic relationships, and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@van_de_pol_identifying_2016]. 
We specified quadratic relationships only because (i) quadratic relationships are more consistent with known biological mechanisms (see Introduction), (ii) preliminary tests revealed that quadratic fits usually had lower AIC, and when not there tended to be little difference in AIC and the curve would approximate a straight line, (iii) modeling only quadratic improves coding efficiency and speeds up the analysis process.
For each group of candidate climate variables, we move forward with the best variable over the time window identified by climwin as a candidate climate variable for the multivariate models.

@@

**The most influential climate drivers...[what tended to be selected?].**
- Time windows
  - often current growing season
  - some exceptions
- Selection within groups:
  - precip vs n precip days
  - temperature group: mean, min, max or PET?

![**Figure 3 | Climate senstivity for all sites.** Columns include the top variables in the precipitation and temperature variable groups. For each species, relationships are plotted if included in top model. Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD. (IN-PREP FIGURE.) ](tables_figures/pre_temp_cld_groups.png)

@@

**Non-linear responses to climate variables were common (Fig. 3).**
- Growth almost always increased with precipitation up to the long-term mean, but often declined under high precipitation.
- The most common response to temperature was unimodal, often peaking within 1SD of long-term mean. (**based on early results-- need to re-evaluate**)

@@

![**Figure 4 | Growth sensitivity to DBH: (a) $\Delta r$, (b) $BAI$, (c) $\Delta AGB$.** Only relationships included in the top model for each site are shown. It's tough to fit all the plots, so we may have to adjust this somehow.](tables_figures/DBH_responses.png)
